Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-05-22 cordillera-cycle.tex
Minor corrections in the intro.

% Authors
\author[1,2]{J.~Seguinot}
\runningauthor{J.~Seguinot et~al.}
\correspondence{J.~Seguinot (julien.seguinot@natgeo.su.se)}

@@

% Affiliations
\affil[1]{Department of Physical Geography and Quaternary Geology and the
          Bolin Centre for Climate Research, Stockholm University,
          Stockholm, Sweden}
\affil[2]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences,
          Potsdam, Germany}

@@

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera have been more extensive than today. At the peak of glaciation, a
contiguous blanket of ice extended from Denali Peak in the north to the Puget
Sound in the south. This blanket is known as the former Cordilleran ice sheet.

@@

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last
glacial cycle. However, former ice thickness, and in turn the ice sheet's
contribution to the last glacial maximum sea-level low-stand, remains poorly
constrained. Moreover, our understanding of the deglaciation phase is
restricted to a regional level, while little is known about the ice sheet
history preceding the last glacial maximum extent. Although time-evolving,
whole ice-sheet reconstructions of glacial advance and retreat patters are
available for the Laurentide and Eurasian ice sheet, this is not the case for
the Cordilleran ice sheet.

The present study aims to use a numerical ice sheet model calibrated against
field-based evidence to attempt a quantitative reconstruction of the
Cordilleran ice sheet history through the last glacial cycle. Although
numerical modelling has been identified as a tool to improve our understanding
of Cordilleran ice sheet more than twenty years ago, the ubiquitously
mountainous topography of the region has presented a major challenge to its
application, which is only overcome by recent development of numerical ice
sheet models and underlying scientific tools. In addition to the topographic
complexity, strong climatic contrasts characteristic of the North American
Cordillera require the use of high-resolution temperature and precipitation
forcing as input data to the ice sheet model.

Here we use the Parallel Ice Sheet Model (PISM), forced by a high-resolution
climatology from the North American Regional Reanalysis, to model the
Cordilleran ice sheet through the last glacial cycle. Because past climate
is mostly unknown, palaeoclimate forcing is applied in the simplified form of
six time-dependent temperature offset time-series. These time-series were
obtained by scaling six different palaeo-temperature reconstructions from
proxy records around the globe, including two \chem{\delta^{18}O} records from
Greenland ice cores, two \chem{\delta^{18}O} records from Antarctic ice cores,
and two alkenone unsaturation index records from Northwest Pacific oceanic
sediment cores. The output from the model is compared to geomorphological
evidence in terms of timing of glaciation, extent of glaciation and patterns
of deglaciation.

@@

Fig.~\ref{fig:locmap} about here.

@@

@@

SSA+SIA, pseudo-plastic sliding, temperature-dependent flow.

\subsection{Bedrock response}

Thermal model and deformation model.

\subsection{Surface mass balance}

Positive degree-day model including spatially and seasonally variable standard
deviation (Fig.~\ref{fig:atm}).

@@

Input climatology from North American Regional Reanalysis according to
sensitivity study from the previous paper (Fig.~\ref{fig:atm}).

@@

\subsection{Palaeo-climate sensitivity test}

All six simulations consistently produce two major glaciations during MIS~4 and
MIS~2 (Figs.~\ref{fig:timeseries} and~\ref{fig:snapshots}).
However the timing is much different. Notably the oceanic core forcing result
in much too early last glacial maximum.

\subsection{High-resolution simulations}

We re-run the GRIP (?) and Vostok (?) core simulations at higher resolution.
Results are
consistent with lower resolution but differences appear due to small-scale
topography better capturing mass-balance lapse-rate effects. In general the
Greenland-based simulation show much more variability.

@@

% References
%\input{references.bbl}
%\newpage

@@

@@

% fig:atm
\begin{figure}
  \includegraphics{atm}
  \caption{Focring climatology. In construction.
           \idea{Show seasonal variability}}
  \label{fig:atm}
\end{figure}

% fig:timeseries
\begin{figure}
  \includegraphics{timeseries}
  \caption{Multi-record timeseries. In construction.
           \todo{Add NGRIP}
           \idea{Use same colors but different patterns for similar records}}
  \label{fig:timeseries}
\end{figure}

% fig:snapshots
\begin{figure}
  \includegraphics{snapshots}
  \caption{Multi-record snapshots. In construction.
           \todo{Add NGRIP}}
  \label{fig:snapshots}
\end{figure}

% fig:duration
\begin{figure}
  \includegraphics{duration}
  \caption{Ice cover duration. In construction.
           \todo{Add Antarctic-record high-resolution simulation}}
  \label{fig:duration}
\end{figure}

% fig:deglac
\begin{figure}
  \includegraphics{deglac}
  \caption{Deglaciation age. In construction.
           \todo{Add Antarctic-record high-resolution simulation}
           \idea{Compare to Dyke deglaciation outlines}}
  \label{fig:deglac}
\end{figure}

>>>>

% Authors
\author[1,2]{J.~Seguinot}
\author[3]{M.~Margold}
\author[2]{I.~Rogozhina}
\author[1]{A.~Stroeven}
\runningauthor{J.~Seguinot et~al.}
\correspondence{J.~Seguinot (julien.seguinot@natgeo.su.se)}

@@

% Affiliations
\affil[1]{Department of Physical Geography and Quaternary Geology and the
          Bolin Centre for Climate Research, Stockholm University,
          Stockholm, Sweden}
\affil[2]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences,
          Potsdam, Germany}
\affil[3]{Department of Geography, Durham University, UK}

@@

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera have been more extensive than today. At the peak of glaciation, a
contiguous blanket of ice extended from Denali Peak in the north to the Puget
Sound in the south. This ice mass is known as the former Cordilleran ice sheet
(Fig.~\ref{fig:locmap}).

@@

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last
glacial cycle. However, former ice thickness, and in turn the ice sheet's
contribution to the last glacial maximum sea-level low-stand, remains poorly
constrained. Moreover, our understanding of the deglaciation phase is
restricted to a regional level, while little is known about the ice sheet
history preceding the last glacial maximum extent. Although time-evolving,
whole ice-sheet reconstructions of glacial advance and retreat patters are
available for the Laurentide and Eurasian ice sheet, such is not the case for
the Cordilleran ice sheet.

The present study aims to use a numerical ice sheet model calibrated against
field-based evidence to perform a quantitative reconstruction of the
Cordilleran ice sheet history through the last glacial cycle. Although
numerical modelling has been identified as a tool to improve our understanding
of Cordilleran ice sheet more than twenty years ago, the ubiquitously
mountainous topography of the region has presented a major challenge to its
application, which is only overcome by recent development of numerical ice
sheet models and underlying scientific tools. In addition to the topographic
complexity, strong climatic contrasts characteristic of the North American
Cordillera require the use of high-resolution temperature and precipitation
forcing as input data to the ice sheet model.

Here we use the Parallel Ice Sheet Model (PISM), forced by a high-resolution
climatology from the North American Regional Reanalysis (NARR), to model the
Cordilleran ice sheet evolution through the last glacial cycle. Because past climate
is mostly unknown, palaeoclimate forcing is simplistically applied as
six time-dependent temperature offset time-series. These time-series were
obtained by scaling six different palaeo-temperature reconstructions from
proxy records around the globe, including two \chem{\delta^{18}O} records from
Greenland ice cores, two \chem{\delta^{18}O} records from Antarctic ice cores,
and two alkenone unsaturation index records from Northwest Pacific oceanic
sediment cores. Model output is compared to geomorphological evidence in terms
of timing of glaciation, extent of glaciation and patterns of deglaciation.

@@

@@

The simulations presented here were run using Parallel Ice Sheet Model (PISM,
development version~11b0a7f), an open-source, finite-difference, shallow ice
sheet model. The model takes basal topography, sea level, geothermal heat flux
and climate forcing, as input fields and computes the evolution of ice extent
and thickness in time, its thermal and dynamic state, and
the associated lithospheric response. Our modelling domain encompasses the
entire area covered by the Cordilleran ice sheet at the last glacial maximum.
It includes independent ice growing on the western Alaskan ranges and the
Brooks Range and intermediate expanses of ice-free terrain in northern Yukon
Territory and interior Alaska (Fig.~\ref{fig:locmap}).

To reconstruct the successive phases growth and decay of the last Cordilleran
ice sheet, palaeo-climatic conditions of the last glacial cycle are mimicked
by applying time-dependent temperature offsets derived from multiple
palaeo-temperature proxy records. Each simulation starts from assumed ice-free
conditions at -120\,kyr, and runs to present. These computations were
performed on 16 to 128 cores at the Swedish National Supercomputing
Center.

@@

The central part of an ice sheet model consists of the computation of flow
velocity which itself depends on temperature. PISM is a~shallow model, which
implies that the balance of stresses is approximated based on their predominant
components. On the other hand, the model is polythermal: it accounts for
differences in temperature and softness within the ice column.

The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
SIA and SSA velocities are computed by finite difference methods on a~10\,km
resolution horizontal grid of 300 by 150 points (the modelling domain). Ice
softness depends on temperature and water content through an enthalpy
formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is
computed in three dimensions in up to 51 irregularly spaced layers within the
ice, and temperature is additionally computed in 31 regularly spaced layers in
bedrock to a~depth of 3\,km. Surface air temperature from the climate forcing
provides the upper boundary condition to the ice enthalpy model, and
a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary
condition to the bedrock thermal model. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial
variations in the geothermal heat flux may significantly affect basal
conditions \citep{pattyn-2010}, this study focuses on reconstructing the former
extent of the Cordilleran ice sheet, on which it has relatively little effect
\citep{rogozhina-etal-2012}.

\todo{Describe the new sliding law?}

A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is computed once at the beginning of the run,
as a~function of modern bed elevation, with lowest values occurring at low
elevations,

\begin{align}
  \phi =
  \begin{cases}
    15      & \text{for }       z<  0 \\
    z/15+15 & \text{for }    0 <z<200 \\
    45      & \text{for }  200 <z     \\
  \end{cases}
\end{align}

where $\phi$ is given in degrees and $z$ in meters above current sea level.
This model thus accounts for a~weakening of till associated with the presence
of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Because the
distribution of marine sediments is assumed constant throughout the run, the
till friction angle $\phi$ is not affected by changes in bed elevation. Basal
topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography
and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.

Sea level is lowered as a function of time according to the SPECMAP
reconstruction, and basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{lingle-clark-1985,bueler-etal-2007}. It uses a lithosphere
density of 3300\,\unit{kg\,m^{-3}}, a flexural rigidity of $5 \times
10^{24}$\,\unit{N\,m} and a mantle viscosity of $1 \times
10^{21}$\,\unit{Pa\,s}. Due to the high mantle viscosity, there exists a time
lag between ice sheet growth and isostatic bedrock response.

\subsection{Surface mass-balance}

Ice surface accumulation and ablation are computed from monthly mean surface
air temperature and monthly precipitation by a~temperature-index (positive
degree-day) model \citep{hock-2003}. Ice accumulation is equal to precipitation
when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
with temperature between 0 and 2\,\unit{{\degree}C}. Ice ablation is computed
from the number of positive degree-days, defined as the integral of
temperatures above 0\,\unit{{\degree}C} in one year.

The positive degree-day integral \citep{calov-greve-2005} is numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology and was
computed from daily temperature values from the North American Regional
Reanalysis \citep{data:narr} and excludes variability associated to the annual
temperature cycle itself \citet{seguinot-rogozhina-2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Atmospheric forcing of the model consists of a present-day monthly climatology
on the one hand, and time-dependent temperature offsets on the other hand. The
present-day climatology was computed from the near-surface air temperature and
precipitation rate fields of the North American Regional Reanalysis (NARR) over
the period xxxx--xxxx.

Prior to surface mass balance computation, the model dynamically applies
a~lapse-rate correction to surface air temperature. This correction accounts
for the evolution of ice thickness on the one hand, and differences between the
climate forcing reference topography and the ice flow model basal topography on
the other hand. It uses a~reference topography distinct from the modelled basal
topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all
simulations. No corrections apply to precipitation changes with elevation.

@@

All six simulations consistently produce two major glaciations during MIS~4 and
MIS~2 (Figs.~\ref{fig:timeseries} and~\ref{fig:snapshots}). However the timing
is much different. Notably the oceanic core forcing result in much too early
last glacial maximum. In general the Greenland-based simulation show much more
variability.

@@

% References
%\input{references.bbl}
\newpage

@@

% tab:records
\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters used to
           prepare temperature offset time-series used to force the ice sheet
           model through the last 120\,\unit{kyr}. $T_{[32;22]}$ refers to the
           mean temperature anomaly during the period -32 to~-22~\unit{kyr} after
           scaling.}
  \label{tab:records}
  {\begin{tabular}{lcccc}
    \tophline
    Record & Proxy & Scaling factor & $T_{[32;22]}$ & Source\\
    \middlehline
    GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C & \aref \\
    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.0{\degree}C & \aref \\
    EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\
    Vostok   & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\
    ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C & \aref \\
    ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C & \aref \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}

@@

% fig:atm
\begin{figure}
  \includegraphics{atm}
  \caption{Monthly mean near-surface air temperature, precipitation and
           standard deviation of daily mean temperature for January and July
           months from the North American Regional Reanalysis (NARR)
           climatology, used as input fields to the ice sheet model. Note the
           strong contrasts in seasonality, timing of the precipitation peak,
           and temperature variability over the model domain, notably between
           the maritime and continental regions.}
  \label{fig:atm}
\end{figure}

% fig:timeseries
\begin{figure}
  \includegraphics{timeseries}
  \caption{\textbf{(top)} temperature offset time-series from ice and sediment
           core records used as palaeo-climate forcing for the ice sheet model
           (Table~\ref{tab:records}). \textbf{(bottom)} Modelled ice volume
           through the last 120\,\unit{kyr}, expressed in meters of sea-level
           equivalent. The grey spans indicate the ranges of modelled times of
           glacial extrema corresponding to MIS~4 (61.9--55.4\,\unit{kyr}), 3
           (52.2--45.6\,\unit{kyr}), and 2 (last glacial maximum,
           29.5--16.9\,\unit{kyr}).}
  \label{fig:timeseries}
\end{figure}

% fig:snapshots
\begin{figure}
  \includegraphics{snapshots}
  \caption{Snapshots of modelled ice surface topography (1\,\unit{km} contours)
           and velocity (\unit{m\,yr^{-1}}) at modelled times of glacial
           extrema, corresponding to MIS~4, 3, and 2.}
  \label{fig:snapshots}
\end{figure}

% fig:duration
\begin{figure}
  \includegraphics{duration}
  \caption{Modelled duration of ice cover during the last 120\,\unit{kyr}.
           Aside the location of present-day ice caps, note the persistence of
           a central ice dome over the Skeena mountains during most of the
           simulation. On the other hand, the maximal extent of the ice sheet
           corresponds to short durations of ice cover along most of the margin.}
  \label{fig:duration}
\end{figure}

% fig:deglac
\begin{figure}
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
           advance exceeded the last glacial maximum advanced are marked in
           green. Hatches denote the Younger Dryas re-advance, which is most
           significant in the GRIP-driven simulation.}
  \label{fig:deglac}
\end{figure}
